---
title: "spatialplots_checkpoint1"
output: html_document
date: "2024-04-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = TRUE, warning = FALSE, message = FALSE}
library(tidyverse)
library(modelr)
library(OpenStreetMap)

# Load 2021 data
df <- read_csv("alleghenycounty_crashdata_2022.csv")
```

# Sample of how to use OpenStreetMap
Taken from https://ajsmit.github.io/Intro_R_Official/mapping-google.html#openstreetmap 

```{r}
library(OpenStreetMap)

# other 'type' options are "osm", "maptoolkit-topo", "bing", "stamen-toner",
# "stamen-watercolor", "esri", "esri-topo", "nps", "apple-iphoto", "skobbler";
# play around with 'zoom' to see what happens; 10 seems just right to me
sa_map <- openmap(c(40.464070, -79.931711), c(40.454649, -79.911975), 
                  type = "esri-topo")

# reproject onto WGS84
sa_map2 <- openproj(sa_map)

# use instead of 'ggplot()'
sa_map2_plt <- OpenStreetMap::autoplot.OpenStreetMap(sa_map2) + 
  xlab("Longitude (째E)") + ylab("Latitude (째S)")
sa_map2_plt
```

Note: 
Coordinates are defined as c(latitude, longitude)
UpperLeft coordinates of a bounding box has most positive latitude and least positive longitude
LowerRight coordinates of a bounding box has the least positive latitude and most positive longitude

# Function for plotting filtered accidents on OpenStreetMap
```{r}
library(OpenStreetMap)

plotAccidentsOSM <- function(df, plotTitle){
  df <- df %>%
    filter(!is.na(DEC_LONG)) %>%
    filter(!is.na(DEC_LAT))
  
  upperLeft <- c(max(df$DEC_LAT), min(df$DEC_LONG))
  lowerRight <- c(min(df$DEC_LAT), max(df$DEC_LONG))
  
  sa_map <- openmap(upperLeft, lowerRight, 
                    type = "esri-topo")
  
  sa_map2 <- openproj(sa_map)
  
  sa_map2_plt <- autoplot(sa_map2) + 
    xlab("Longitude (째E)") + ylab("Latitude (째S)") + 
    ggtitle(plotTitle) + 
    geom_point(data = df,
               aes(x = DEC_LONG, y = DEC_LAT), # slightly shift the points
               colour = "red", size =  1)
  
  sa_map2_plt
}
```


# ROAD_CONDITION = Slush Roads
```{r}
slushRoads <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(ROAD_CONDITION == 4)

nrow(slushRoads)
```

# ROAD_CONDITION = Snow covered
```{r}
snowRoads <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(ROAD_CONDITION == 3)

nrow(slushRoads)
```

# ILLUMINATION = Dark - no street lights
```{r}
darkNoStreetLights <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(ILLUMINATION == 2)

plotAccidentsOSM(darkNoStreetLights,"FATAL_OR_MAJ_INJ, ILLUMINATION=Dark - no street lights")
```

```{r}
darkNoStreetLights_INJ_FATAL <- df %>%
  filter(INJURY_OR_FATAL == 1) %>%
  filter(ILLUMINATION == 2)

plotAccidentsOSM(darkNoStreetLights_INJ_FATAL,"INJURY_OR_FATAL, ILLUMINATION=Dark - no street lights")
```

```{r}
darkNoStreetLights_INJ_FATAL <- df %>%
  filter(INJURY_OR_FATAL == 1) %>%
  filter(ILLUMINATION == 2) %>%
  count(STREET_NAME) %>%
  arrange(desc(n))

darkNoStreetLights_INJ_FATAL
```

```{r}
darkNoStreetLights_PENN_LINCOLN_PY_INJ_FATAL <- df %>%
  filter(INJURY_OR_FATAL == 1) %>%
  filter(ILLUMINATION == 2) %>%
  filter(STREET_NAME == "PENN LINCOLN PY")

plotAccidentsOSM(darkNoStreetLights_PENN_LINCOLN_PY_INJ_FATAL, "INJURY_OR_FATAL, ILLUMINATION=Dark - no street lights,\n Street=PENN LINCOLN PY")
```


```{r}
darkNoStreetLights_RAYMOND_HW_INJ_FATAL <- df %>%
  filter(INJURY_OR_FATAL == 1) %>%
  filter(ILLUMINATION == 2) %>%
  filter(STREET_NAME == "RAYMOND P SHAFER HW")

plotAccidentsOSM(darkNoStreetLights_RAYMOND_HW_INJ_FATAL, "INJURY_OR_FATAL, ILLUMINATION=Dark - no street lights,\n Street=RAYMOND P SHAFER HW")
```

# ILLUMINATION = Dark with street lights
```{r}
darkStreetLights <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(ILLUMINATION == 3)

plotAccidentsOSM(darkStreetLights, "FATAL_OR_MAJ_INJ, ILLUMINATION=Dark with street lights")
```

```{r}
darkStreetLights <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(ILLUMINATION == 3) %>%
  count(STREET_NAME) %>%
  arrange(desc(n))

darkStreetLights
```
```{r}
darkStreetLights_PennLinconlnPy <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(ILLUMINATION == 3) %>%
  filter(STREET_NAME == "PENN LINCOLN PY")

plotAccidentsOSM(darkStreetLights_PennLinconlnPy, "FATAL_OR_MAJ_INJ, ILLUMINATION=Dark with street lights,\n Street=PENN LINCOLN PY")
```

```{r}
darkStreetLights_PennLinconlnPy_INJ_FATAL <- df %>%
  filter(INJURY_OR_FATAL == 1) %>%
  filter(ILLUMINATION == 3) %>%
  filter(STREET_NAME == "PENN LINCOLN PY")

plotAccidentsOSM(darkStreetLights_PennLinconlnPy_INJ_FATAL, "INJURY_OR_FATAL, ILLUMINATION = Dark with street lights,\n Street = PENN LINCOLN PY")
```

```{r}
darkStreetLights_zoomIn <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(ILLUMINATION == 3) %>%
  filter(DEC_LAT < 40.5 & DEC_LAT > 40.3) %>%
  filter(DEC_LONG < -79.8 & DEC_LONG > -80.1)

plotAccidentsOSM(darkStreetLights_zoomIn, "FATAL_OR_MAJ_INJ, ILLUMINATION=Dark with street lights (Zoom In)")
```

# CURVED_ROAD = 1
```{r}
curvedRoad <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(CURVED_ROAD == 1)

plotAccidentsOSM(curvedRoad, "FATAL_OR_MAJ_INJ, CURVED_ROAD = 1")
```

```{r}
curvedRoad <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(CURVED_ROAD == 1) %>%
  count(STREET_NAME) %>%
  arrange(desc(n))

curvedRoad
```


```{r}
curvedRoad_PENN <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(CURVED_ROAD == 1) %>%
  filter(STREET_NAME == "PENN LINCOLN PY")

plotAccidentsOSM(curvedRoad_PENN, "FATAL_OR_MAJ_INJ, CURVED_ROAD = 1,\n STREET=PENN LINCOLN PY")
```

```{r}
curvedRoad_BABCOCK <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(CURVED_ROAD == 1) %>%
  filter(STREET_NAME == "BABCOCK BL")

plotAccidentsOSM(curvedRoad_BABCOCK, "FATAL_OR_MAJ_INJ, CURVED_ROAD = 1,\n STREET=BABCOCK BL")
```

```{r}
curvedRoad_FIFTH <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(CURVED_ROAD == 1) %>%
  filter(STREET_NAME == "FIFTH AV")

plotAccidentsOSM(curvedRoad_FIFTH, "FATAL_OR_MAJ_INJ, CURVED_ROAD = 1,\n STREET=FIFTH AV")
```

# COLLISION_TYPE = Hit pedestrian
```{r}
hitPedestrian <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(COLLISION_TYPE == 8)

plotAccidentsOSM(hitPedestrian, "FATAL_OR_MAJ_INJ, COLLISION_TYPE = Hit pedestrian")
```

```{r}
hitPedestrian_zoom <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(COLLISION_TYPE == 8) %>%
  filter(DEC_LAT < 40.5 & DEC_LAT > 40.4) %>%
  filter(DEC_LONG < -79.9 & DEC_LONG > -80.05)

plotAccidentsOSM(hitPedestrian_zoom, "FATAL_OR_MAJ_INJ, COLLISION_TYPE = Hit pedestrian (Zoom In)")
```

# COLLISION_TYPE = Rear-end
```{r}
rearEndCollision <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(COLLISION_TYPE == 1)

plotAccidentsOSM(rearEndCollision, "FATAL_OR_MAJ_INJ, COLLISION_TYPE = Rear-end")
```

# LOCATION_TYPE = Ramp and bridge
```{r}
rampAndBridge <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(LOCATION_TYPE == 8)

nrow(rampAndBridge)
```

# LOCATION_TYPE Bridge
```{r}
bridge <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(LOCATION_TYPE == 3)

nrow(bridge)
```

# MUNICIPALITY = 02449
```{r}
municipality02449 <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(MUNICIPALITY == '02449')

nrow(municipality02449)
```

# INTERSECT_TYPE = Y intersection
```{r}
yIntersection <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(INTERSECT_TYPE == '03')

plotAccidentsOSM(yIntersection, "FATAL_OR_MAJ_INJ, INTERSECT_TYPE = Y intersection")
```

# ROAD_OWNER = County road
```{r}
countyRoad <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(ROAD_OWNER == 3)

plotAccidentsOSM(countyRoad, "FATAL_OR_MAJ_INJ, ROAD_OWNER = County road")
```

# ROAD_OWNER = State Highway
```{r}
stateHighway <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(ROAD_OWNER == 2)

plotAccidentsOSM(stateHighway, "FATAL_OR_MAJ_INJ, ROAD_OWNER = State Highway")
```


# IMPAIRED_DRIVER = 1
```{r}
impairedDriver <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(IMPAIRED_DRIVER == 1)

plotAccidentsOSM(impairedDriver, "FATAL_OR_MAJ_INJ, IMPAIRED_DRIVER = 1")
```

# UNBELTED = 1
```{r}
unbelted <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(UNBELTED == 1)

plotAccidentsOSM(unbelted, "FATAL_OR_MAJ_INJ, UNBELTED = 1")
```

# AGGRESSIVE_DRIVING = 1
```{r}
aggressiveDriving <- df %>%
  filter(FATAL_OR_MAJ_INJ == 1) %>%
  filter(AGGRESSIVE_DRIVING == 1)

plotAccidentsOSM(aggressiveDriving, "FATAL_OR_MAJ_INJ, AGGRESSIVE_DRIVING = 1")
```
